---
title: "Computer Setup for Code Club"
toc: true
date-modified: 2023-08-18
---

```{r knitr_options, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

----

<br>

Here, you will find general information about computer setup for Code Club.
Additional setup instructions for individual sessions will appear in the posts for each session.

In summary, you just need to get **access to R and RStudio**,
preferably via a local installation,
and optionally via the Ohio Supercomputer Center (OSC).

If you **run into issues or have questions**,
don't hesitate to contact one of the organizers.
You can also come to Code Club 15 minutes early,
and one or more of the organizers should be there already.

<br>

## Install R {#install-r}

- **Windows**: Download and run the `.exe` file for the latest version of R from <https://cran.r-project.org/bin/windows/base/>,
  by clicking the large `Download R [version-number] for Windows` link at the top of the gray box.

- **Mac**: Download and run the `.pkg` file for the latest version of R from <https://cran.r-project.org/bin/macosx/>,
  by clicking the link just below `Latest release`.

::: {.callout-note}
### Already have R installed?

- Please check your version of R -- this information is printed to the console
  when you start R, and you can also get it by typing `sessionInfo()` and checking
  the first line of the output.
  
- Currently (August 2023), we would recommend R version `4.2.0` or higher.
  And since we'll work with the "base R pipe",
  you'll definitely need version `4.1.0` or higher.
  
- To update R, [see this page for instructions](/tutorials/r-resources-tips/#updating-r).

:::  

## Install RStudio {#install-rstudio}

RStudio is a so-called Integrated Development Environment (IDE) for R,
with side-by-side panes for an R script, an R concole, plots, help documents, and much more.
While it is perfectly possible to use R without RStudio, RStudio has become
the de facto standard for working with R and is very useful.

To install RStudio, go to the [RStudio download page](https://rstudio.com/products/rstudio/download/#download)
and download and run the installer file for your operating system. 

## Quarto-website specific installation

For the Fall 2023 sessions on creating a Quarto website,
you will need to have the _rmarkdown_ R package installed.
You can do so by typing (or copying) the following into your R Console inside
RStudio:

```{r}
install.packages("rmarkdown")
```

While not necessary for basic website building inside RStudio,
it will also be a good idea to **install Quarto** itself.
This is not an R package, but is a separate piece of software.
To install it,
go to <https://quarto.org/docs/get-started/> and the correct installer
for your operating system should be highlighted in a blue box.
Click on that box to download and once downloaded,
click on the installer to install Quarto.

<br>

## Alternative: Use RStudio Server at OSC {#osc}

Upon request ([contact Jelmer](mailto:poelstra.1@osu.edu)),
you can get access to the [Ohio Supercomputer Center (OSC)](http://osc.edu)
Classroom Project for Code Club (`PAS1838`).
This way, you can code in RStudio from your browser rather than with a local installation.
This is a good option if you prefer not to install anything or if you run
into problems during installations.

**After you asked for access to the OSC project,
you should receive an email from OSC that you have been added to the Code Club OSC project.**

- *If you already have an OSC account*,
  you shouldn't need to do anything to gain access,
  although the email may ask you to confirm/accept your being added to project.

- *If you do not yet have an OSC account*,
  the email you received from OSC should have a link to do so.
  Alternatively, follow the instructions below to sign up and get access to the project.

::: {.callout-tip collapse="true"}

#### Run RStudio Server from the OSC OnDemand website (Click to expand) {#osc-run-rstudio}

[OSC OnDemand](https://ondemand.osc.edu/) lets you access OSC resources through your
browser and run applications like RStudio.

1. To get started, go to <https://ondemand.osc.edu/> and log in with your OSC username and password.

2. Click on "Interactive Apps" in the blue top bar, and select "RStudio Server" (near the bottom).
  
3. Now, you're on a page from which you can launch an RStudio server that will run on an OSC cluster.
   Select project PAS1838 in the dropdown menu and change the "Number of hours" to `2`.
   Then click "Launch".

4. You will be sent to a page where you can see the status of your "job"
   It usually starts running within seconds, and the color of the top
   bar will then switch from blue ("Queued" and then "Starting") to green ("Running").
   
5. Click "Connect to RStudio Server" at the bottom of the box,
   and an RStudio Server instance will open.

:::

<br>

## More info {#more}

Please see our **[R Resources and Tips](./r_resources.qmd)** 
if you are new to R or need to update your R version!
