---
title: "Making maps in R: part II"
subtitle: "..."
pagetitle: "Maps II"
author: "Jelmer Poelstra"
date: "2025-02-24"
categories: [maps]
title-block-banner: false
number-depth: 2
execute: 
  eval: false
  cache: false
knitr:
  opts_chunk:
    out.width: "85%"
editor_options: 
  chunk_output_type: console
---

------------------------------------------------------------------------

We will also use some data sets from the _usdata_ (as in: US data) package,
so go ahead and install that:

```{r, eval=FALSE}
install.packages("usdata")
```

With both packages installed, we can now load them:

```{r}
library(usdata)
library(tidyverse)
```

## Bonus content: plotting Ohio cities

We'll need two additional packages:

```{r, eval = FALSE}
install.packages("maps")
install.packages("ggrepel")
```

```{r}
library(maps)
library(ggrepel)
```

```{r}
oh_cities <- us.cities |>
  filter(country.etc == "OH") |>
  mutate(name = sub(" OH", "", name))
```

```{r}
ggplot(ohio) +
  geom_polygon(aes(x = long, y = lat, group = group),
               fill = "grey90", color = "grey70") +
  geom_point(data = oh_cities,
             aes(x = long, y = lat, size = pop),
             fill = "steelblue3", shape = 21) +
  geom_text(data = oh_cities, aes(x = long, y = lat - 0.1, label = name)) +
  coord_sf()
```

```{r}
ggplot(ohio) +
  geom_polygon(aes(x = long, y = lat, group = group),
               fill = "grey90", color = "grey70") +
  geom_point(data = oh_cities,
             aes(x = long, y = lat, size = pop),
             fill = "steelblue3", shape = 21) +
  geom_text_repel(data = oh_cities, aes(x = long, y = lat, label = name)) +
  coord_sf()
```

## A first choropleth map

In choropleth maps, the color of a geographic area differs based on some variable,
like population density, median income, mean temperature, or biodiversity.
In ggplot-speak, we would be mapping the `fill` aesthetic to a variable like
population density.

```{r}
states |>
  filter(region %in% c("ohio", "indiana", "michigan")) |> 
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group, fill = region),
               color = "grey40", linewidth = 0.5) +
  coord_sf()
```

<hr style="height:1pt; visibility:hidden;" />

```{r}
state_stats
```

```{r}
state_stats_ed <- state_stats |>
  mutate(region = tolower(state))

states_with_stats <- tibble(states) |>
  left_join(state_stats_ed, by = "region")
```

```{r}
ggplot() +
  geom_polygon(
    data = states_with_stats,
    aes(x = long, y = lat, group = group, fill = med_income / 1000),
    color = "black", linewidth = 0.1,
  ) +
  coord_sf(crs = 5070, default_crs = 4326) +
  scale_fill_viridis_c(name = "Median income\n(thousands) ")
```


## Maps not colored by predefined areas

## Plotting routes

